Speculation in foreign exchange: noise or information

### 2015-06-10 Knitr/NewPaper.Rnw
Updated the whole NewPaper Now it needs to be read through and corrected.

\documentclass[12pt, a4paper, oneside]{article} % Paper size, default font size and one-sided paper
%\graphicspath{{./Figures/}} % Specifies the directory where pictures are stored
%\usepackage[dcucite]{harvard}
\usepackage{rotating}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{pdflscape}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage[comma, sort&compress]{natbib}% Use the natbib reference package - read up on this to edit the reference style; if you want text (e.g. Smith et al., 2012) for the in-text references (instead of numbers), remove 'numbers' 
\usepackage{graphicx}
%\bibliographystyle{plainnat}
\bibliographystyle{agsm}
\usepackage[colorlinks = true, citecolor = blue, linkcolor = blue]{hyperref}
%\hypersetup{urlcolor=blue, colorlinks=true} % Colors hyperlinks in blue - change to black if annoying
%\renewcommand[\harvardurl]{URL: \url}
 \usepackage{listings}
  \usepackage{tikz}
 \usetikzlibrary{arrows,positioning}
 \usepackage{color}
 %\graphicspath{{../Pictures/}}
\definecolor{mygrey}{gray}{0.95}
\lstset{backgroundcolor=\color{mygrey}}
\begin{document}
\title{Speculation in foreign exchange: noise or information}
\author{Rob Hayward\footnote{University of Brighton Business School, Lewes Road, Brighton, BN2 4AT; Telephone 01273 642586.  rh49@brighton.ac.uk. }} 
\date{\today}
\maketitle
\begin{abstract}
An event study is used to assess the Keynesian and Friedman views of speculation. Speculative extremes are ranked by intensity of sentiment and weight of activity.  The extreme says little about the future in either case, suggesting that speculation is more than just random noise and supporting the view there is some informational content that is being passed through to the price with speculative activity. A unique dataset of risk-reversal skew on option prices is used to measure the intensity of speculative sentiment; the weight of positions is based on US regulatory data.  
\end{abstract}
\section*{Introduction}
There are a range of ideas about the way that speculation can affect financial markets.  Towards one end of this spectrum is the view of Keynes from Chapter 13 of \emph{The General Theory} where he identified speculation as a myopic, sentiment-driven activity, dominated by the desire to ``beat the gun'' or  ``outwit the crowd'' \citet[p. 101]{Keynes1936}. Here speculative activity is a problem that, at its most extreme, will cause booms, reversals and will discourage fundamental, long-term investment. It is noise that does not emerge randomly, but feeds off itself as a collective sentiment  encourages  trends and serial correlation in asset prices.

@@

The aim of this paper is to understand more about the nature of speculation by investigating the relationship between both the intensity of speculation and the weight of speculative positions with the movement of prices in the foreign exchange market.   If speculators are uninformed and their activity can be regarded as noise, the most extreme sentiment and the time when speculators have the greatest weight in the market should correspond with the greatest deviation of prices from fundamental value, when reversals are most likely to take place; if speculators are informed prices should follow a random walk after either of these extremes.  
    The foreign exchange market is utilised because it is the largest and most liquid market in the world.  The latest survey taken by the Bank for International Settlements (BIS) in April 2013 estimated average daily turnover at \$5.4trn with average daily spot transactions of \$2.0trn \citet{BISFX2013}. In addition, the cost of trading in this markket, when measurd in terms of the size of the bid-ask spread, is extremely low. An analysis of the bid-ask spreads for USD-JPY, GBP-USD and EUR-USD for the period January 2001 to 2005 and finds that the average spread for each is 0.342\%, 0.305\% and 0.326\% respectively.  This would be the cost for a round-trip (buying and selling immediately)\citep{Steely2013}.

@@

\section{Literature}
\subsection*{Overview}
  Speculation is an important concept in economics, particular in the area of finance.  However, there is no clear, unambiguous theory of speculation. In a number of fields the term \emph{noise-trading} has been used for something like the activity discussed by Keynes and Fridman.   \citet{bloomfield2009noise} suggests that there are three categories of noise-trader: uninformed speculators, market-making liquidity-providers and informed speculators. The first of these groups is similar to the picture of speculation provided by Keynes, the second refers to those agents or institutions paid for providing liquidity and for taking the risk of meeting an informed trader in the market (these are also called market-makers), and the final category is close to the idea of speculation that has been presented by Friedman. 
  % is this the place to state that market-making will not be considered separately?  %this paragraph must go somewhere else.  I do not know where. 
Nonetheless, the distinction between market-making speculative activity is not precise.\footnote{The implementation of the \emph{Volcker} Rule in the US which will limit the scope of US investment banks to trade on their own account has raised the question of where market-making turns into speculation.}

In the \emph{microstructure} literature, the term noise-trader is often interchanged with \emph{liquidity-trader}.  They do not usually distinguish between psychological and institutional reasons for the lack of attention to fundamental information. Nonetheless, liquidity tends to refer to institutional explanations for trading without reference to fundamental value. This may mean that they just execute customer orders at the current price or it may be that they buy and sell with reference to a benchmark index or to fulfil a hedging strategy.  Examples of the microstructure approach are found in \citet{Kyle1985Continuous} and \citet{glosten1985bid}. See \citep{Lyons2006Microstructure} and \citep{Evans2002Order} for examples of the micrstrucure approach in foreign exchange.

The other issue is whether uninformed speculative noise-trading, of either variety, can be considered as just random noise?   The original microstrucutre version had moise-traders arriving in a random fashion so that the market-maker could infer information from an imbalance of orders \citet{Kyle1985Continuous}.  However, though the idea that noise is a random variable may be a construction that is useful for modelling and it may be justified in analysis of broader themes of microstructure and levels of market efficiency, this exogenous construct may be an over-simplification that is inapprpropriate when looking for more understanding of speculation itself.  Institutional and psychological explanations for noise-trading frequently suggest that there are endogenous non-random elements.
% is this the right places for this.  If so, it must be expanded. If not, it can go somewhere near the interaction noise and informaed traders. That is line 58 currently and mark a.  
The modelling of noise is related to the rise of rational expectations in economics.

\subsection*{The problem of uninformed noise-traders}
The traditional Keynesian view of speculation asserted that it would be an additional noise that would make it even harder for informed-traders to overcome uncertainty and apply fundamental information about long-term value. This view finds some supported in Shiller's observation that stock prices are more volatile than the underlying earnings information on which valuation should be based \citet{ShillerVol}. In addition, \citet{Delong1990noise} and \citet{shleifer1997limits} show that constrained horizons can cause noise-trading to limit the ability of informed traders to take advantage of profitable opportunities for fear of further divergence from fundamental value before realignment is achieved. 
%It would be good to have some more modern ideas here.  Maybe they are just the empirical results. 
\subsection{Interplay between Informed and uninformed}
\citet[p. 1]{GortonNoise} categorise noise-traders as all those trading in markets for ``non-information-based reasons''. In this case, the construction olves the problem of financial market trading. If economic agents know the fundamental value of a security how can some be buying and some be selling? There must be different opinions.  Why Will traders be prepared to pay the cost of acquiring information if they are unable to take advantage of that information? However, with noise-traders, the informed can take advantage of their information.  This creaes a justification for acquiring information. Noise-traders can therefore create a gap between the price and the fundamtental value that will encourage informed traders to pay the cost of acquiring information. The divergence between this price and fundametnal value is essential to encourage te process of price discovery. 
% What where the reasons for creating the concept of noise-traders?  There are three:  the rtionale for trading, the motivtion or mraket-marking and... 
Fisher Black identified noise as ``the arbitrary element in expectations'' \citet[p.529]{BlackNoise}. Therefore, noise makes financial markets possible but imperfect: the more noise-traders that there are, the more liquidity and the easier it is to trade; the more noise-traders, the higher the level of inefficiency in the market and the more likely that price will be driven from fundamental value.

\subsection{Other empirical work}
There is little empirical eassessment of these theories about speculation and noise-trading. In part this is because information is limited and it is usually difficult to distinguish between speculators and other investors and, even more so, between informed and uninformed.

One approach has been to assume that noise-traders utilise simple investment strategies that involve momentum or reversals.  For example, \citet{ShillerFashion} discusses the use of fashionable techniques or fads in investment.  In this way, speculation noise is associated with sentiment and measures of sentiment are then used as ways of measuring speculation. Alternatively, speculation can be identified with momentum trading.   The literature on heterogeneous traders, another way that the trading securities is explained, has very frequently used noise-traders as trend-chasers or followers of technical analysis.  These models (see for example brock and hommes [1998]; lux [1995]; lux and marchesi [2000]) investigate bubbles and crashes, and feature fundamental traders versus noise traders.

@@

The paper makes use of the utility function to compare the outcomes for the two types of investor for different realisations of the parameters.  They note four effects:  the \emph{Friedman Effect}, which is the tendency of informed investors to push prices back towards fundamental value; \emph{Price Pressure} as the way that speculative noise-traders buy more and drive the price higher as they become more bullish; \emph{Hold More},whereby the noise-traders hold more of the risky asset; \emph{Create Space} is the added risk that is caused by an increase in the variability of noise-trader belief.  The Friedman Effect and Price Pressure tend to reduce noise-trader utility as they work to create losses or reduced gains for noise-trader positions; Hold More and Create Space can increase noise-trader utility and profits \citep[pp. 14-15]{Delong1990noise}. 
%Put something on these four in the conclusion. 
A number of important ideas flow from the model.  The main findings are that noise-traders will not necessarily disappear as Friedman had suggested \citep{FriedmanPositive}.  While noise-traders will tend to get less utility than sophisticated traders, this does not necessarily drive them from the market.  The noise-traders expected to get higher utility due to their misperception.  The reduction in utility comes mainly from the fact that they are not compensated for taking risk that they are not aware of.  This is consistent with the type of over-confidence that has been found in day-traders \citet{odean1998volume}

De Long and the others test a variety of specifications that include modest switching from one style of investment to another and changes in the weight of new traders after previous performance.  Some result in the disappearance of noise-traders while others suggest that they can become more dominant.  The model also explains the importance of investment with a long-term horizon as this is the only way that \emph{noise- trader risk} can be overcome.  If agents live for more than two periods, fundamental investors can make an arbitrage between safe and risky assets, driving the price back to fundamental value.  Therefore, investors or traders with a reputation that will allow resources and time to consider the long-term would be able to take full advantage of the noise-traders' misperception.  The model relies crucially on the institutional or psychological myopia. 
% read again and add more.

Noise-traders drive the price in the direction of their mis-perception.    The second and third terms in Equation \ref{eq:NP2} show the deviation from fundamental value due to the fact that the noise-trader mis-perception is not equal to zero but equal to $p^*$.  This means that the mis-perception is not random but biased in a positive fashion.  If the mis-perception is random, we are back to the random walk with noise-traders adding liquidity on each side of the market and the informed traders able to take advantage of any deviation of the price from its fundamental value.  There is still some risk that the outcome will be different from the fundamental value, because the noise-traders may still become more extreme in their misperception, but on average informed traders will make profit if they act.  This deviation in the price is what De Long and co-authors call price pressure.  While there is no fundamental risk in this model, price pressure causes risk for which there is no compensation and which is assumed by the noise-traders.  This is rather similar to the risk created by speculators in the Keynes models where additional uncertainty, in the Knightian sense, encourages more investors towards self-perpetuating myopic behaviour.

>>>>

\documentclass[12pt, a4paper, oneside]{article} % Paper size, default font size and one-sided paper
%\graphicspath{{./Figures/}} % Specifies the directory where pictures are stored
%\usepackage[dcucite]{harvard}
\usepackage{rotating}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage{pdflscape}
\usepackage[flushleft]{threeparttable}
\usepackage{multirow}
\usepackage[comma, sort&compress]{natbib}% Use the natbib reference package - read up on this to edit the reference style; if you want text (e.g. Smith et al., 2012) for the in-text references (instead of numbers), remove 'numbers' 
\usepackage{graphicx}
%\bibliographystyle{plainnat}
\bibliographystyle{agsm}
\usepackage[colorlinks = true, citecolor = blue, linkcolor = blue]{hyperref}
%\hypersetup{urlcolor=blue, colorlinks=true} % Colors hyperlinks in blue - change to black if annoying
%\renewcommand[\harvardurl]{URL: \url}
 \usepackage{listings}
  \usepackage{tikz}
 \usetikzlibrary{arrows,positioning}
 \usepackage{color}
 %\graphicspath{{../Pictures/}}
\definecolor{mygrey}{gray}{0.95}
\lstset{backgroundcolor=\color{mygrey}}
\begin{document}
\title{Speculation in foreign exchange: noise or information}
\author{Rob Hayward\footnote{University of Brighton Business School, Lewes Road, Brighton, BN2 4AT; Telephone 01273 642586.  rh49@brighton.ac.uk. }} 
\date{\today}
\maketitle
\begin{abstract}
An event study is used to assess the views of Keynes and Friedman on speculation. Speculative extremes are ranked by intensity of sentiment and weight of speculative activity. A unique dataset of risk-reversal skew on option prices is used to measure the intensity of speculative sentiment; the weight of positions is based on US regulatory data.  The extremes say little about the future in either case, suggesting that speculation is more than just random noise and supporting the view there is some informational content that is being passed through to the price with speculative activity. 
\end{abstract}
\section*{Introduction}
There are a range of ideas about the way that speculation can affect financial markets.  Towards one end of this spectrum is the view of Keynes from Chapter 13 of \emph{The General Theory} where he identified speculation as a myopic, sentiment-driven activity, dominated by the desire to ``beat the gun'' or  ``outwit the crowd'' \citet[p. 101]{Keynes1936}. Here speculative activity is a problem that, at its most extreme, will cause booms, reversals and will discourage fundamental, long-term investment. It is noise that does not emerge randomly, but feeds off itself as the evolution of collective sentiment encourages trends in asset prices.

@@

The aim of this paper is to understand more about the nature of speculation by investigating the relationship between both the intensity of speculation and the weight of speculative positions with the movement of prices in the foreign exchange market.   If speculators are uninformed and their activity can be regarded as noise, the most extreme sentiment and the time when speculators have the greatest weight in the market should correspond with the greatest deviation of prices from fundamental value, when reversals are most likely to take place; if speculators are informed prices should follow a random walk after either of these extremes.  
    The foreign exchange market is utilised because it is the largest and most liquid market in the world.  The latest survey taken by the Bank for International Settlements (BIS) in April 2013 estimated average daily turnover at \$5.4trn with average daily spot transactions of \$2.0trn \citet{BISFX2013}.  The cost of trading in this markket, when measurd in terms of the size of the bid-ask spread, is extremely low. An analysis of the bid-ask spreads for USD-JPY, GBP-USD and EUR-USD for the period January 2001 to 2005 and finds that the average spread for each is 0.342\%, 0.305\% and 0.326\% respectively.  This would be the cost for a round-trip (buying and selling immediately)\citep{Steely2013}.

@@

\section{Literature}
\subsection*{Overview}
  Speculation is an important concept in economics, particular in the area of finance.  However, there is no clear, unambiguous theory of speculation. Speculation plays a key role in three importantant and related areas: market-microstructure, the rationale for the trading of financial securities and rational expectations.

One key element of the microstructure literature involves distinguishing between uninformed or noise-traders and those who have infomration. The Kyle model has market-makers reciving orders from the informed and the uninformed.  On the assumption that the uniformed arrive in a random fashion without any reason to trade, the weight of orders implies something about the activity of the informed and causes prices to change in the process of price discovery.  The bid-ask spread is the payment that the market-maker recieves in return for taking the risk of meeting an informed trader.  Ultimately, the uninformed traders make payments to the informed traders with the mark-makers taking a cut for the risk and the privision of exchange \citet{Kyle1985Continuous} and \citet{glosten1985bid}.  Here the speculators may be informed or uninformed.

% does this following paragraph come later?  I think so....Once the three
% main categories are done. 
In the \emph{microstructure} literature, the term noise-trader is often interchanged with \emph{liquidity-trader}.  They do not usually distinguish between psychological and institutional reasons for the lack of attention to fundamental information. Nonetheless, liquidity tends to refer to institutional explanations for trading without reference to fundamental value. This may mean that they just execute customer orders at the current price or it may be that they buy and sell with reference to a benchmark index or to fulfil a hedging strategy.

\citet[p. 1]{GortonNoise} categorise noise-traders as all those trading in markets for ``non-information-based reasons''. In this case, uninforrmed speculation provides a rational and an opportunity for the trading of financial securities. If all economic agents know the fundamental value of a security how can some be buying and some be selling? There must be different opinions.  Why Will traders be prepared to pay the cost of acquiring information if they are unable to take advantage of that information? With speculative noise-traders, the informed can take advantage of their information.  This creates a justification for acquiring information. Speulative noise-traders can therefore create a gap between the price and the fundamtental value that will encourage informed traders to pay the cost of acquiring information. The divergence between this price and fundametnal value is essential to encourage te process of price discovery. 
%
Fisher Black suggested that this combination was essential to the smooth function of financial markets. It is ``the arbitrary element in expectations'' \citet[p.529]{BlackNoise}. Therefore, noise makes financial markets possible but imperfect: the more noise-traders that there are, the more liquidity and the easier it is to trade; the more noise-traders, the higher the level of inefficiency in the market and the more likely that price will be driven from fundamental value.

  Speculation is also an important part of the argument for why prices should tend towards that consistent with \emph{rational expectations}.  \citet{muth1961} uses commodity speculation to argue for model-consistent expectation formation.  In this case, the speculators are informed and tend to reduce the volatility of prices.

% See \citep{Lyons2006Microstructure} and \citep{Evans2002Order} for examples of the micrstrucure approach in foreign exchange.

If there are a mixture of informed and uninformed specualtors, the question becomes whether the uninformed speculative activity can be considered as just random noise?   The original microstrucutre version had moise-traders arriving in a random fashion so that the market-maker could infer information from an imbalance of orders \citet{Kyle1985Continuous}.  However, though the idea that noise is a random variable may be a construction that is useful for modelling and it may be justified in analysis of broader themes of microstructure and levels of market efficiency, this exogenous construct may be an over-simplification that is inapprpropriate when looking for more understanding of speculation itself.  Institutional and psychological explanations for noise-trading frequently suggest that there are endogenous non-random elements.

\subsection*{The problem of uninformed noise-traders}
The traditional Keynesian view of speculation asserted that it would be an additional noise that would make it even harder for informed-traders to overcome uncertainty and apply fundamental information about long-term value. This view finds some supported in Shiller's observation that stock prices are more volatile than the underlying earnings information on which valuation should be based \citet{ShillerVol}. In addition, \citet{Delong1990noise} and \citet{shleifer1997limits} show that constrained horizons can cause noise-trading to limit the ability of informed traders to take advantage of profitable opportunities for fear of further divergence from fundamental value before realignment is achieved. 
%It would be good to have some more modern ideas here.  Maybe they are just the empirical results. 
%this paragraph is not very good.  It needs to be updated.  Look at the file in citulike. 
\subsection{Other empirical work}
There is little empirical eassessment of these theories about speculation and noise-trading. In part this is because information is limited and it is usually difficult to distinguish between speculators and other investors and, even more so, between informed and uninformed.

One approach has been to assume that noise-traders utilise simple investment strategies that involve momentum or reversals.  For example, \citet{  shillerFashion} discusses the use of fashionable techniques or fads in investment.  In this way, speculation noise is associated with sentiment and measures of sentiment are then used as ways of measuring speculation. Alternatively, speculation can be identified with momentum trading.   The literature on heterogeneous traders, another way that the trading securities is explained, has very frequently used noise-traders as trend-chasers or followers of technical analysis.  These models (see for example brock and hommes [1998]; lux [1995]; lux and marchesi [2000]) investigate bubbles and crashes, and feature fundamental traders versus noise traders.

@@

The following effects are noted:  the \emph{Friedman Effect}, which is the tendency of informed investors to push prices back towards fundamental value; \emph{Price Pressure} as the way that speculative noise-traders buy more and drive the price higher as they become more bullish; \emph{Hold More},whereby the noise-traders hold more of the risky asset; \emph{Create Space} is the added risk that is caused by an increase in the variability of noise-trader belief \citep[pp. 14-15]{Delong1990noise}. 
%Put something on these four in the conclusion. 
The main findings are that noise-traders will not necessarily disappear as Friedman had suggested \citep{FriedmanPositive}.  While noise-traders will tend to get less utility than sophisticated traders, this does not necessarily drive them from the market.  The noise-traders expected to get higher utility due to their misperception.  The reduction in utility comes mainly from the fact that they are not compensated for taking risk that they are not aware of.  This is consistent with the type of over-confidence that has been found in day-traders \citet{odean1998volume}

\subsection{The implications of the noise-trader model}
The aim of this study is to test the implications of the noise-trader model.  There are two parameters from the noise-trader price Equation \ref{eq:NP2} that can be used to measure the amount of speculation:  the price misperception $(\rho)$ and the weight of noise-traders in the market $(\mu)$.  The aim of this section is to  identify the returns that are implied by the noise-trader model when either the intensity of the misperception or the weight of noise-traders is increased to extreme levels so that this can be used as a benchmark against which foreign exchange market activity can be judged.

To understand what the noise-trader model of speculative behaviour means for foreign exchange rates, a \emph{Monte Carlo} simulation is carried out with a variety of mis-perceptions and noise-trader weights.  Two time series are created.  The first series is based on a random noise-trade misperception, $(\rho^*)$.  This is called the Misperception Model (MPM). The second is based on a random weight of noise-traders $(\mu)$ and is called the Random Weight Model (RWM).   The simulation creates a 1000 period price series from Equation \ref{eq:NP2}. The dividend paid on the risky asset $(r)$ is posted at 10\%; the risk-aversion parameter $(\gamma)$ is assumed to be 0.3 (in line with that suggested in \citep[p. 127]{FrobesBehaviour}); starting from a price of 100, the average noise-trader mis-perception $(p^*)$ is a random variable in the first simulation with a mean of 4 and a variance $(\sigma_{\rho^*}^2)$ of 2, and has a fixed value of 4 in the second simulation; the weight of noise-traders $(\mu)$ is 50\% in the first case but is modelled as a two-period moving average of a normal random variable centred at 50\% with a standard deviation of 20\% and equally weighted average of its previous value (starting at 50\%) to allow smooth adjustment, in the second simulation.

A 1000 period prices simulation is created for each model and the returns for these series are calculated as the log change in price.  Those periods when mis-perception or the weight of speculators is \emph{extreme} are identified as those that are at or above the ninety-fifth percentile.  These extremes are the \emph{event} for an event study. The package \emph{eventstudies} \citep{eventstudies} aligns the defined events (in this case the extreme noise-trader activity) so that descriptive statistics and other analysis can be carried out on the period after the event has occurred.  Descriptive statistics for the whole series of returns and for those when misperception or weight of noise-traders are presented in Table \ref{tabref:MC1}.  The process is then repeated for one thousand simulations of each model so that confidence intervals can be established for the estimates of the descriptive statistics.

\subsection{The results of simulation}
The results of the simulation are recorded in Table \ref{tabref:MC1}.  It shows that the mean return to an investment strategy in the whole of the sample is very close to zero.  However, for each of the extreme readings, there are large negative returns to follow. This is consistent with the structure of the model: noise-traders drive price above intrinsic value and subsequent returns will be negative; the more extreme the misperception or the weight of uninformed speculative noise-traders in the market, the greater the negative return that is likely to be recorded.

\begin{table}[h!]
\begin{threeparttable}
\caption{Results of the Monte Carlo Simulation of the Noise-Trader Model}
\begin{tabular}{p{6cm} | r r r r}
 \label{tabref:MC1}
                            & Base Case  & Extreme (MPM) & Extreme (RWM) \\
  \hline
\hline
Mean                     & -0.00 & -0.90*   & -0.49 \\
Standard Deviation  & 0.98 &  0.49*     &  0.42*\\
Skew                      &  -0.07 &  -0.16   &  -0.78*\\
Kurtosis                  &  -0.14 &  -0.38    &  -0.35\\
Number                 &  1000        &   50        & 50\\

\end{tabular}
\begin{tablenotes}
\small
\item Descriptive statistics for the De Long Noise-Trader model \citep{Delong1990noise}.  The Base Case measures the benchmark as returns from prices generated by Equation \ref{eq:NP2} for the whole sample.   The MPM measures returns after a period of extreme sentiment when the misperception of noise-traders ($\rho$) is equal to or above the 95th percentile of all the misperceptions; the RMW measures the returns following a recording of extreme weight of noise-traders ($\mu$) is equal to or above the 95th percentile of all noise-trader realisations.   The asterisk denotes significantly different from zero at the 95th confidence interval.    
\end{tablenotes}
\end{threeparttable}
\end{table}

% this needs to be completed. 
\subsection{Measuring the intensity of speculative sentiment}
The measurement of speculative intensity must identify how strongly speculators believe that the market will move in a particular direction. In the absence of a daily survey of speculators, this study uses \emph{risk-reversal skew} as an objective measure of speculative sentiment rather than surveys of opinion.\footnote{There are a number of surveys of investor opinion that are regularly conducted by financial service information companies.  However, these are disparate and, in many cases, unsystematic, and it is not clear whether they measure speculative sentiment or investor opinion more generally.   The Wall Street Journal runs an annual survey of investor opinion, but this is focused mostly on US equities.  The \emph{Economist} magazine also solicits opinion, but in a relatively unstructured and \emph{ad hoc} fashion.  In the foreign exchange market, there is a broad set of market forecasts that was collected by the financial services company \emph{MMS International} (formerly Money Market Services) which polled a broad range of financial market participants on key exchange rates for one and four week forecast horizons. Frankel and Froot found that the median forecast from the MMS survey exhibited trend-following behaviour at the one to three week horizon while there is mean-reversion beyond that point.  Frankel and Froot also assess the 'Financial Report' that has been produced by \emph{The Economist} magazine and \emph{Currency Forecasters Digest} and find a wide dispersion of opinion and express some concern about the aggregation of such divergent forecasters \citep{FrootFrankelFDB} and \citep{FrankelFroot1987} \label{fnsurvey}.}    This is the relative price of currency put and call options for the same strike price.  The \emph{Put-Call-Parity} in the Black-Scholes option pricing model assumes that these will be the same \citep{Black1973Options}.  While the Black-Scholes model is the basis for valuing options, where the price of a non-dividend paying stock trades at a discount to the forward rate equal to the risk free rate, the forward rate in a currency option is determined by the relative interest rate differential between the two currencies.   The Garman-Kohlhagan adjustment that accounts for the difference in interest payments on the two currencies is used.   The spot discount is determined therefore by the relative interest rate differential for the appropriate maturity \citep{GarmanKohlhagan}.   This model allows the market practice of pricing options from the volatility implied by the model.  In other words, the Garman-Kohlhagan is used to extract the volatility that is implied by a specific option price given the strike and other parameters.

Determining the fundamental value of exchange rates has proved to be extremely difficult.  As PPP is, at best, a medium-term phenomenon and given the finding by Meese and Rogoff about the superiority of the random walk relative to other economic models and accepting for the moment the EMH, it will be assumed that the forward rate is the best estimate of the future value of the exchange rate.  Any forecast which deviates from the forward rate will in this case be counted as a misperception and the further away from the random walk the greater the misperception. \footnote{The expected  exchange rate should be equal on average to the forward rate once any risk premia has been take into account. Though it will be seen that UIP and the assumption that the future exchange rate is on average equal to the interest rate differential does not hold, the fact that daily data is being used here and the attention is on the most extreme divergences from UIP suggests that this will not greatly affect the results.}

This study uses a unique dataset of \emph{risk-reversals} for eleven major currency pairs.  This is the only period for which the data currently exist.  The risk-reversal data is very difficult to obtain in an over-the-counter market like foreign exchange where there is no centralised exchange to record and collect information, generally being the property of quoting institutions.  The risk-reversals are the average prices quoted at 10:00 GMT by major banks on the Thomson-Reuters composite option pages (OPT1)  for the period between January 2 1996 and September 6 2002.  The risk-reversals are the spread of implied volatility for a 1 month 25 delta at-the-money call option relative to the equivalent put.  A call option is the right to buy and a put option is the right to sell.  The delta of an option measures the effect of a change in the spot price on the value of the option.  This can range from zero for deeply out-of-the-money options to unity for deeply in the money strikes.  The at-the-money delta is 50 as there is a 50:50 chance of the option being exercised.  The benchmark for option risk-reversals is 25 delta.  This is a measure of how far the option is out of the money or how far the strike is from the forward rate.

If the exchange rate trajectory is expected to follow a random walk with drift towards the 1 month forward, implied volatility on the put and the call for the same delta should be equal.  Any deviation from this equivalence is an indication that a deviation from the random walk with drift is implied in the market price of options.\footnote{If the risk-reversal for a 1-month USD/CAD is quoted at 0.15-0.28\% and the implied volatility is 8.50\%, the market-maker would be willing to buy the 1-month 25 delta USD-put-CAD-call at 8.65\% and sell the USD-call-CAD-put at 8.50\%.  The dealer pays 0.15\%.  Alternatively, the dealer would sell a 25 delta call at 8.78\% and buy the USD-call-CAD-put at 8.50\%, earning 0.28\% \citep{Global}. }  Therefore, the relationship between implied volatility for calls and implied volatilities for puts for the same delta and maturity can be taken as an indication of how much more expensive calls are relative to the equivalent put or as the market bias towards puts over calls or, in this case, as an indication of market sentiment.  The larger the absolute level of the risk-reversal skew, the greater the relative price and, in equilibrium, the greater the belief, expressed through the action of market participants or the inventory of market-makers, that there will be a deviation from the forward rate.   If extreme deviations or speculations are noise they should presage a sharp reversal.  If speculation is informed, this is part of the price discovery process and extremes may be followed by calm conditions around a new equilibrium.

% The table may be pushed onwards.  Use t (i think). 
\begin{landscape}
\begin{table}[t]
\begin{threeparttable}
\caption{Descriptive Statistics for Daily Exchange Rate Returns: 2 January 1996 to 6 September 2002}
\begin{tabular}{l | r r  r r r r r r r r}
  \hline
 Statistic &  AUD & EURGBP & EURSEK & GBP & CHF & EURCHF & EURJPY & EUR & CAD & JPY \\
  \hline
Number & 1685 & 1675 & 1716 & 1667 & 1674 & 1674 & 1675 & 1668 & 1683 & 1675  \\
Mean & -0.0174 & -0.0187 & 0.0011 & 0.0025 & 0.0120 & -0.0039 & -0.0072 & -0.0172 & 0.0084 & 0.0076\\
Median &  0.0000 & -0.0298 & 0.0000& 0.0000 & 0.0114 & 0.0000 & 0.0000 & -0.0253 & 0.0000 & 0.0188\\
Standard deviation  & 0.7241 & 0.5542 & 0.0438 & 0.0489 & 0.6819 & 0.0277 & 0.8416 & 0.6338 & 0.0363 & 0.8170\\
Max & 5.8301 & 3.0543 & 2.1633 & 2.2573 & 3.5086 & 1.6477 & 4.0879 & 2.4136 & 2.2170 & 5.1788 \\
Min & -3.4666 & -2.2858 & -2.2900 & -2.2701 & -3.6525 & -1.8466 & -8.2848 & -2.2901 & -2.9707 & -10.6573 \\
Skewness  & 0.1206* & 0.2923* & 0.0432 & -0.0452 & -0.1423* & -0.4046* & -0.7388* & 0.1738* & -0.1095 & -1.7237*\\
Kurtosis & 4.17* & 1.88 & 2.43* & 1.48 & 1.96 & 5.84* & 7.08* & 1.12 & 5.11* & 21.52*\\
\hline
 \label{tabref:DS1}
\end{tabular}
\begin{tablenotes}
\small
\item Continuously compounded returns (multiplied by 100) for period 2 January 1996 to  September 2002 for US dollar per unit of Australian dollar (AUD), UK sterling per unit of Euro (EURGBP), Swedish Crown per unit of EUR (EURSEK); US dollar per unit of UK sterling (GBP);  Swiss franc per unit of US dollar (CHF); Swiss franc per unit of Euro (EURCHF); US dollars per unit of Euro (EUR); Canadian dollars per unit of US dollar (CAD); Japanese yen per unit of US dollars (JPY). The asterisk (*) means that the estimate is significantly different from zero at the 5\% level using t-statistics for the mean and the standard error of skew and standard error of excess-kurtosis respectively.     
\end{tablenotes}
\end{threeparttable}
\end{table}
\end{landscape}

Table \ref{tabref:DS1} shows descriptive statistics for the exchange rate data used in the study.  There are daily data for the period between January 1996 and September 2002.  This is the period for which risk-reversal data are available.  The series are of sightly different lengths as there are some gaps in the risk reversal and exchange rate data as a result of holidays and market closures.  From the raw data continuously compounded returns are calculated and multiplied by 100.  The first thing of note is that the mean return does not appear to be statistically different from zero.  This is what would be expected if, as has been identified in many studies, exchange rates follow a random walk. This apparent stability hides some significant changes that are taking place below the surface.  For example, the Japanese yen records an appreciation of 10.66\% on one day.  Most of the exchange rate series appear to be skewed in some way and there is also evidence of kurtosis in many cases.  An asterisk is used to denote statistical significance at the level of 95\% confidence.  A t-test is used for the mean; a \emph{standard error of skew} and a \emph{standard error of kurtosis} are used for the skew and kurtosis respectively.  These are calculated as $SES = \sqrt{\frac{6n(n-1)}{(n-2)(N+1)(N+3)}}$ and $SEK = \sqrt{\frac{n^2-1}{(n-3)(N+5)}}$, with the assumption that this estimate is distributed normally, the skew or kurtosis is considered significant if it is more than two standard errors from zero.

\begin{landscape}
\begin{table}[t!]
\begin{threeparttable}
\caption{Descriptive Statistics for Risk Reversals: 2 January 1996 to 6 September 2002}
\begin{tabular}{l | r r  r r r r r r r r}
  \hline
 Statistic &  AUD & EURGBP & EURSEK & GBP & CHF & EURCHF & EURJPY & EUR & CAD & JPY \\
 \hline
Number & 1702 & 1697 & 1719 & 1691 & 1374 & 1339 & 1697 & 1692 & 1701 & 1697 \\ 
  Mean & -0.37 & 0.15 & 0.71 & -0.08 & -0.28 & -0.38 & -0.48 & 0.19 & 0.12 & -0.74 \\ 
  Median & -0.30 & 0.10 & 0.80 & -0.10 & -0.15 & -0.30 & -0.10 & 0.10 & 0.15 & -0.55 \\ 
  Standard Deviation & 0.38 & 0.37 & 0.51 & 0.33 & 0.49 & 0.36 & 1.02 & 0.49 & 0.25 & 1.02 \\ 
 Max & 0.45 & 1.72 & 1.90 & 1.00 & 1.05 & 0.50 & 1.70 & 2.05 & 1.45 & 2.05 \\ 
 Min & -1.70 & -0.80 & -0.30 & -1.35 & -2.05 & -1.80 & -4.03 & -1.15 & -0.65 & -4.00 \\ 
 Skewness & -0.50* & 1.27* & -0.23* & 0.33* & -0.79* & -0.58* & -0.93* & 0.68* & 0.23* & -0.51* \\ 
  Kurtosis & 0.02 & 2.29* & -1.67 & 0.72 & 0.69 & 0.19 & 0.15 & 0.96 & 1.83 & -0.03 \\ 
      \hline
 \label{tabref:DS2}
\end{tabular}
\begin{tablenotes}
\small
\item Daily risk reversal skew.  Risk reversal is the premium of implied volatility for calls relative to the equivalent put (equal distance to the strike price, 25 delta in this case).  A figure above zero means that calls are more expensive (in volatility and price terms) than puts.  A figure below zero means that puts are more expensive. The asterisk (*) means that the estimate is significant at the 5\% level using t-statistics for the mean and the standard error of skew and standard error of excess-kurtosis respectively.   These are calculated as $SES = \sqrt{\frac{6n(n-1)}{(n-2)(N+1)(N+3)}}$ and $SEK = \sqrt{\frac{n^2-1}{(n-3)(N+5)}}$, with the assumption that this estimate is distributed normally, the skew or kurtosis is considered significant if it is more than two standard errors from zero.
\end{tablenotes}
\end{threeparttable}
\end{table}
\end{landscape}
